Type Evolution in Interactive Programming Systems

### 2024-05-30 paper.tex
fix diagram reference
remove tikz diagrams

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{tikz-cd}
\usepackage{float}
\usepackage{wrapfig}

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{0.5em}
\includegraphics[scale=0.25]{figures/diag-divergence.png}
\end{wrapfigure}
\frameworkboxtitle{Type Branching with Divergence}
The divergence scenario adds another dimension to the problem. There are now multiple
versions of the system with different \emph{types} that may co-exist as multiple branches.
Edits to \emph{data} are done against a specific type, but they then need to be migrated
to versions with different types. Consequently, any type evolution needs to be accompanied
both with forward projection (transforming data to a new version) and backward projection
(transforming new data to the old version).}

@@

\newpage
~
\newpage

\section{diagrams}

\begin{figure}[H]
  \centering
  \begin{tikzcd}[sep = 6em]
    \textsf{query}
    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
    &
    \textsf{schema}
    \arrow[l, "\textsf{constrains}"{above}]
    \arrow[r, "\textsf{constrains}"{above}]
    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
    &
    \textsf{data}
    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
    \\
    \textsf{query}_\alpha
    &
    \textsf{schema}_\alpha
    \arrow[l, "\textsf{constrains}"{above}]
    \arrow[r, "\textsf{constrains}"{above}]
    &
    \textsf{data}_\alpha
    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
  \end{tikzcd}
  \caption{Extract Entity}
\end{figure}

\begin{figure}[H]
  \centering
  \begin{tikzcd}[column sep = large, row sep = 5em]
    &
    \textsf{schema}
    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
    \arrow[rr, "\textsf{constrains}"]
    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
    &
    &
    \textsf{data}
    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
    \arrow[Rightarrow, from=L, to=R]
    \\
    \textsf{schema}_\alpha
    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
    \arrow[rr, crossing over]
    &
    &
    \textsf{data}_\alpha
    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
    &
    \\
    &
    \textsf{schema}_\beta
    \arrow[ld, "\textsf{change}_\alpha \succ\, \textsf{change}_\beta"{name=L4, near start, description}]
    \arrow[rr]
    &
    &
    \textsf{data}_\beta
    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
    \arrow[Rightarrow, from=L4, to=R4]
    \\
    \textsf{schema}_{\alpha\beta}
    \arrow[rr]
    &
    &
    \textsf{data}_{\alpha\beta}
    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
    &
  \end{tikzcd}
  \caption{Mergeing divergent changes with rebase ($\succ$)}
\end{figure}

\begin{figure}[H]
  \centering
  \begin{tikzcd}[column sep = large, row sep = 5em]
    &
    \textsf{schema}
    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
    \arrow[rr, "\textsf{constrains}"]
    \arrow[dd, "\textsf{identity}"{near start, description}]
    &
    &
    \textsf{data}
    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
    \\
    \textsf{schema}_\alpha
    \arrow[rr, crossing over]
    &
    &
    \textsf{data}_\alpha
    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
    &
    &
    \textsf{data}_{\alpha\gamma}
    \arrow[dd, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha \succ\, \textsf{update}_\gamma"{near start, description}]
    \\
    &
    \textsf{schema}
    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
    \arrow[rr]
    &
    &
    \textsf{data}_\beta
    \arrow[ld, "\textsf{migrate}_\alpha"{name=R4, near start, description}]
    \arrow[Rightarrow, from=L4, to=R4]
    \\
    \textsf{schema}_\alpha
    \arrow[rr]
    &
    &
    \textsf{data}_{\alpha\beta}
    \arrow[from=uu, crossing over, "\textsf{update}_\beta \succ\, \textsf{migrate}_\alpha"{near start, description}]
    &
    &
    \textsf{data}_{\alpha\beta\gamma}
  \end{tikzcd}
  \caption{Conference Organizer merging schema and divergent data updates}
\end{figure}

\textcolor{red}{[JE] These diagrams are getting hairy. Are they actually helpful?}

>>>>

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{float}
\usepackage{wrapfig}

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{0.5em}
\includegraphics[scale=0.25]{figures/old/diag-divergence.png}
\end{wrapfigure}
\frameworkboxtitle{Type Branching with Divergence}
The divergence scenario adds another dimension to the problem. There are now multiple
versions of the system with different \emph{types} that may co-exist as multiple branches.
Edits to \emph{data} are done against a specific type, but they then need to be migrated
to versions with different types. Consequently, any type evolution needs to be accompanied
both with forward projection (transforming data to a new version) and backward projection
(transforming new data to the old version).}

@@
